### 2019-09-19 submission/manuscript.Rmd
Add more text

At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski and Chemistry guy]. By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on. Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology*, it is worth stepping back and using the development of mothur as a story that likely has parallels to many other research stories that have taken time to develop.

@@

***Competition is good and healthy.***

* A lot of people have been working in this space - e.g. clustering algorithms
* Approaches to developing software have differed (e.g. web-based (greengenes/RDP), boutique (Quince), wrappers (QIIME))

@@

The future
* 16S rRNA gene sequencing is not dead
* Always improving sequencing methods bring new problems
* Larger datasets, need a replacement for naive open and closed reference clustering
* Lingering controversies - singletons, ASVs, contamination

Longterm funding
* The development of mothur was enabled by initial funding from a grant to Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative.
* NSF
* NIH
* Bootstrapping from other projects

Red Queen

>>>>

At the risk of navel gazing, looking back on our scientific journeys can be instructive to other scientists who are overwhelmed at the prospect of looking forward at their careers [Lenski, Chemistry guy, Lyme disease mbio, (A)historical paper]. By no means is my personal journey over, but since 2002 I have been on a journey that I did not realize I was on. Now that the paper introducing the mothur software package is ten years old and has become the most cited paper published by *Applied and Environmental Microbiology*, it is worth stepping back and using the development of mothur as a story that likely has parallels to many other research stories that have taken time to develop.

@@

***Competition is good and healthy.*** From the beginning there have been online tools the Ribosomal Database Project (RDP), greengenes, and SILVA. These allowed users a straightforward method of comparing their data to those collected in a database. There are two primary downsides to these tools. First, researchers running the online tool must pay the computational expenses leading to slow process times as numerous users simultaneously attempt an analysis and when hardware becomes outdated. Eventually this limitation would result in the termination of the greengenes website. Second, these platforms provide a one-size-fits-all analysis. These tools only allow a user to analyze 16S and in some cases 18S rRNA gene sequences. If a users analyzes a different gene then the tool will not serve them. These observations resulted in two design goals we have had with mothur: bringing the analysis to a user's computer and separating a tool from the database. For example, we commonly use a sequence alignment method that was originally developed for greengenes, but use the SILVA databases. In addition, we offer the naive Bayesian classifier developed by the RDP and allow users to train it to any database they want, including customized databases. In both examples, users can align or classify non rRNA gene sequence data. As the bioinformatics tools have matured, both the RDP and SILVA offer pipelines for analyzing large datasets, albeit in one-size-fits-all black box implementations.

With the growth in popularity of 16S rRNA gene sequencing there has naturally been an expansion in the number of people developing tools to analyze these data. Months after the paper describing mothur was released, QIIME was released with similar goals. Over the past 10 years, many have attempted to compare these two programs: Pepsi vs Coke, Apple vs Windows, etc. It is never clear which software is which brand and whether it is a complement or an insult. Regardless, both programs are very popular. From my perspective, most of the differences are cosmetic. QIIME is effectively a bundle of scripts to run other developers' software. For example, with QIIME v1, it was possible to run mothur as part of QIIME. One can also run the naive Bayesian classifier through QIIME using the original code developed by the RDP. From my perspective, this caused QIIME many problems because there were numerous software dependencies that had to be installed causing frustration for users. Although the QIIME developers would create virtual machines and packaging tools to simplify installation, these fixes required sophistication by the user. In contrast, when a user runs mothur, they are running mothur. The naive Bayesian classifier code that is in mothur is a rewritten version of the original code. When we rewrite someone's software we do it with an eye to improving performance, access, and utility for non-16S rRNA gene sequence data. For example, while 454 data was popular, PyroNoise was an effective tool for denoising flowgram data. Running the original code required a linux computer and knowledge of bash and Perl scripting. When we rewrote the code for mothur, we made it accessible to people using any operating system with a simple command interface. I have found that our approach requires significant developer effort, but saves considerable user effort.

Beyond the large packages like mothur and QIIME, there has been significant growth in stand alone boutique software tools for sequence curation (e.g. PyroNoise and DADA), chimera checking (e.g. UCHIME, ChimeraSlayer, and Perseus), and clustering (e.g. USEARCH and Swarm). Where possible, we have implemented these algorithms directly into mothur. We have also used this diversity of methods to perform head-to-head comparisons. Most notable is the area of clustering algorithms where there have been a large number of algorithms developed without an obvious method to objectively compare them. Through a series of studies, we applied an objective metric, the Matthew's Correlation Coefficient (MCC), to numerous algorithms for clustering sequences into OTUs. By performing this type of analysis, we were able to objectively compare the algorithms, make recommendations to the field, and develop new algorithms that outperformed the existing algorithms. We have embraced the competition and diversity of all methods being used to analyze amplicon data as an opportunity to identify the strengths and weaknesses of the methods in an attempt to make recommendations to other researchers.

@@

**The future.** I plan to continue developing mothur for as long as other researchers find it useful. One challenge of such a plan is maintaining the funding to support its development. The development of mothur was initially enabled by a grant to Mitch Sogin from the Sloan Foundation to support his VAMPS (Visualization and Analysis of Microbial Population Structures) initiative. We used that seed funding to secure an NSF grant and then a grant from NIH for tool development as part of their Human Microbiome Project. Since that project expired in 201X, we have not had funding to specifically support mothur's development. I have been fortunate to have discretionary funds generated from other projects to help support mothur. Although there is funding for new tools, there appears to be little appetite by funders to support existing tools. Emblematic of this was the NIH program, Big Data To Knowledge (BD2K), which solicited proposals through the program announcement "Extended Development, Hardening and Dissemination of Technologies in Biomedical Computing, Informatics and Big Data Science (PA-14-156)". This opportunity appeared perfect, except that the National Institute of Allergy and Infectious Diseases (NIAID), the primary supporter of microbiome research at NIH, did not participate in the announcement. Tools like mothur have been successful and need mechanisms to continue to mature and support the needs of the research community.

As with anything in science, methods become passe. When we first developed mothur, tools like T-RFLP and DGGE were still commonly used. Today it would be hard to argue that data from those methods meaningfully advance a study relative to what one could get using 16S rRNA gene sequence data. Looking forward, many want to claim that amplicon sequencing is today's DGGE. They claim that instead researchers should move on to shotgun metagenomic sequence data. It is important to note that the two methods answer fundamentally different questions. Metagenomic sequence data tells a researcher about the functional potential and genetic diversity of a community and 16S rRNA gene sequence data describes the taxonomic composition. Both tools provide important information but they cannot easily replace each other. Although metagenomic data does provide highly resolved taxonomic information, the limit of detection is much higher than that of amplicon data. For example, we analyzed 10,000 16S rRNA sequences from each of about 500 subjects. We can think of this as representing about 1,000,000 genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S rRNA gene sequences/genome). Assuming a genome is 4 Mbp, this would represent a sequencing depth of 4 Tbp. Although such a sequencing effort is possible, albeit expensive, analyzing such a large dataset with an approach that captures the full genetic diversity of the community would be financially and technically prohibitive. Going forward, there is still a place for 16S rRNA gene sequencing. Although sequencing technologies will continue to evolve to capture longer and more high quality data, there will likely always be a need for characterizing the taxonomic diversity of microbial communities. With this in mind, there will always be a place for tools like mothur.

Of course this does not mean that such tools will remain static. We see three key areas that we will continue to help to develop. First, just as we adapted through the transitions from Sanger to 454 to MiSeq sequencing platforms, we must learn how to adapt data from PacBio, Oxford Nanopore, and other tools to our analysis pipelines. As with these earlier platforms, we must understand the error profiles so that they can be corrected. We have learned that moving forward requires that we maintain or improve sequence quality. No doubt, datasets and read lengths will improve. Second, with these improvements, we will need to continue to improve our algorithms. We have already seen that attempts to use low quality MiSeq and HiSeq data causes computational problems leading to the creation of open and closed reference clustering methods. Unfortunately, comparative analyses showed that these methods fail relative to *de novo* clustering methods. More work is needed to improve reference-based clustering methods so that larger datasets can be analyzed without sacrificing quality. Finally, there are ongoing controversies that need further exploration. These include the validity and utility of amplicon sequence variants, the validity and impact of removing low frequency sequences, and methods of identifying and removing 16S rRNA gene sequences from contaminants. In the original mothur paper, we commented that 16S rRNA gene sequencing and analysis is very much like the Red Queen in Alice in Wonderland. Although some disagreed with this analogy, we still feel it fits. The sequencing technology and rapacious appetite of researchers continues to race on. At the same time, bioinformatics tools must adapt to facilitate our research. I am confident that mothur will be up to this exciting challenge.

### 2019-09-19 submission/manuscript.tex
Add more text

At the risk of navel gazing, looking back on our scientific journeys can
be instructive to other scientists who are overwhelmed at the prospect
of looking forward at their careers {[}Lenski and Chemistry guy{]}. By
no means is my personal journey over, but since 2002 I have been on a
journey that I did not realize I was on. Now that the paper introducing
the mothur software package is ten years old and has become the most
cited paper published by \emph{Applied and Environmental Microbiology},
it is worth stepping back and using the development of mothur as a story
that likely has parallels to many other research stories that have taken
time to develop.

@@

\textbf{\emph{Competition is good and healthy.}}

\begin{itemize}
\tightlist
\item
  A lot of people have been working in this space - e.g.~clustering
  algorithms
\item
  Approaches to developing software have differed (e.g.~web-based
  (greengenes/RDP), boutique (Quince), wrappers (QIIME))
\end{itemize}

@@

The future * 16S rRNA gene sequencing is not dead * Always improving
sequencing methods bring new problems * Larger datasets, need a
replacement for naive open and closed reference clustering * Lingering
controversies - singletons, ASVs, contamination

Longterm funding * The development of mothur was enabled by initial
funding from a grant to Sogin from the Sloan Foundation to support his
VAMPS (Visualization and Analysis of Microbial Population Structures)
initiative. * NSF * NIH * Bootstrapping from other projects

Red Queen

>>>>

At the risk of navel gazing, looking back on our scientific journeys can
be instructive to other scientists who are overwhelmed at the prospect
of looking forward at their careers {[}Lenski, Chemistry guy, Lyme
disease mbio, (A)historical paper{]}. By no means is my personal journey
over, but since 2002 I have been on a journey that I did not realize I
was on. Now that the paper introducing the mothur software package is
ten years old and has become the most cited paper published by
\emph{Applied and Environmental Microbiology}, it is worth stepping back
and using the development of mothur as a story that likely has parallels
to many other research stories that have taken time to develop.

@@

\textbf{\emph{Competition is good and healthy.}} From the beginning
there have been online tools the Ribosomal Database Project (RDP),
greengenes, and SILVA. These allowed users a straightforward method of
comparing their data to those collected in a database. There are two
primary downsides to these tools. First, researchers running the online
tool must pay the computational expenses leading to slow process times
as numerous users simultaneously attempt an analysis and when hardware
becomes outdated. Eventually this limitation would result in the
termination of the greengenes website. Second, these platforms provide a
one-size-fits-all analysis. These tools only allow a user to analyze 16S
and in some cases 18S rRNA gene sequences. If a users analyzes a
different gene then the tool will not serve them. These observations
resulted in two design goals we have had with mothur: bringing the
analysis to a user's computer and separating a tool from the database.
For example, we commonly use a sequence alignment method that was
originally developed for greengenes, but use the SILVA databases. In
addition, we offer the naive Bayesian classifier developed by the RDP
and allow users to train it to any database they want, including
customized databases. In both examples, users can align or classify non
rRNA gene sequence data. As the bioinformatics tools have matured, both
the RDP and SILVA offer pipelines for analyzing large datasets, albeit
in one-size-fits-all black box implementations.

With the growth in popularity of 16S rRNA gene sequencing there has
naturally been an expansion in the number of people developing tools to
analyze these data. Months after the paper describing mothur was
released, QIIME was released with similar goals. Over the past 10 years,
many have attempted to compare these two programs: Pepsi vs Coke, Apple
vs Windows, etc. It is never clear which software is which brand and
whether it is a complement or an insult. Regardless, both programs are
very popular. From my perspective, most of the differences are cosmetic.
QIIME is effectively a bundle of scripts to run other developers'
software. For example, with QIIME v1, it was possible to run mothur as
part of QIIME. One can also run the naive Bayesian classifier through
QIIME using the original code developed by the RDP. From my perspective,
this caused QIIME many problems because there were numerous software
dependencies that had to be installed causing frustration for users.
Although the QIIME developers would create virtual machines and
packaging tools to simplify installation, these fixes required
sophistication by the user. In contrast, when a user runs mothur, they
are running mothur. The naive Bayesian classifier code that is in mothur
is a rewritten version of the original code. When we rewrite someone's
software we do it with an eye to improving performance, access, and
utility for non-16S rRNA gene sequence data. For example, while 454 data
was popular, PyroNoise was an effective tool for denoising flowgram
data. Running the original code required a linux computer and knowledge
of bash and Perl scripting. When we rewrote the code for mothur, we made
it accessible to people using any operating system with a simple command
interface. I have found that our approach requires significant developer
effort, but saves considerable user effort.

Beyond the large packages like mothur and QIIME, there has been
significant growth in stand alone boutique software tools for sequence
curation (e.g.~PyroNoise and DADA), chimera checking (e.g.~UCHIME,
ChimeraSlayer, and Perseus), and clustering (e.g.~USEARCH and Swarm).
Where possible, we have implemented these algorithms directly into
mothur. We have also used this diversity of methods to perform
head-to-head comparisons. Most notable is the area of clustering
algorithms where there have been a large number of algorithms developed
without an obvious method to objectively compare them. Through a series
of studies, we applied an objective metric, the Matthew's Correlation
Coefficient (MCC), to numerous algorithms for clustering sequences into
OTUs. By performing this type of analysis, we were able to objectively
compare the algorithms, make recommendations to the field, and develop
new algorithms that outperformed the existing algorithms. We have
embraced the competition and diversity of all methods being used to
analyze amplicon data as an opportunity to identify the strengths and
weaknesses of the methods in an attempt to make recommendations to other
researchers.

@@

\textbf{The future.} I plan to continue developing mothur for as long as
other researchers find it useful. One challenge of such a plan is
maintaining the funding to support its development. The development of
mothur was initially enabled by a grant to Mitch Sogin from the Sloan
Foundation to support his VAMPS (Visualization and Analysis of Microbial
Population Structures) initiative. We used that seed funding to secure
an NSF grant and then a grant from NIH for tool development as part of
their Human Microbiome Project. Since that project expired in 201X, we
have not had funding to specifically support mothur's development. I
have been fortunate to have discretionary funds generated from other
projects to help support mothur. Although there is funding for new
tools, there appears to be little appetite by funders to support
existing tools. Emblematic of this was the NIH program, Big Data To
Knowledge (BD2K), which solicited proposals through the program
announcement ``Extended Development, Hardening and Dissemination of
Technologies in Biomedical Computing, Informatics and Big Data Science
(PA-14-156)''. This opportunity appeared perfect, except that the
National Institute of Allergy and Infectious Diseases (NIAID), the
primary supporter of microbiome research at NIH, did not participate in
the announcement. Tools like mothur have been successful and need
mechanisms to continue to mature and support the needs of the research
community.

As with anything in science, methods become passe. When we first
developed mothur, tools like T-RFLP and DGGE were still commonly used.
Today it would be hard to argue that data from those methods
meaningfully advance a study relative to what one could get using 16S
rRNA gene sequence data. Looking forward, many want to claim that
amplicon sequencing is today's DGGE. They claim that instead researchers
should move on to shotgun metagenomic sequence data. It is important to
note that the two methods answer fundamentally different questions.
Metagenomic sequence data tells a researcher about the functional
potential and genetic diversity of a community and 16S rRNA gene
sequence data describes the taxonomic composition. Both tools provide
important information but they cannot easily replace each other.
Although metagenomic data does provide highly resolved taxonomic
information, the limit of detection is much higher than that of amplicon
data. For example, we analyzed 10,000 16S rRNA sequences from each of
about 500 subjects. We can think of this as representing about 1,000,000
genome equivalents (10,000 16S rRNA genes/subject x 500 subjects / 5 16S
rRNA gene sequences/genome). Assuming a genome is 4 Mbp, this would
represent a sequencing depth of 4 Tbp. Although such a sequencing effort
is possible, albeit expensive, analyzing such a large dataset with an
approach that captures the full genetic diversity of the community would
be financially and technically prohibitive. Going forward, there is
still a place for 16S rRNA gene sequencing. Although sequencing
technologies will continue to evolve to capture longer and more high
quality data, there will likely always be a need for characterizing the
taxonomic diversity of microbial communities. With this in mind, there
will always be a place for tools like mothur.

Of course this does not mean that such tools will remain static. We see
three key areas that we will continue to help to develop. First, just as
we adapted through the transitions from Sanger to 454 to MiSeq
sequencing platforms, we must learn how to adapt data from PacBio,
Oxford Nanopore, and other tools to our analysis pipelines. As with
these earlier platforms, we must understand the error profiles so that
they can be corrected. We have learned that moving forward requires that
we maintain or improve sequence quality. No doubt, datasets and read
lengths will improve. Second, with these improvements, we will need to
continue to improve our algorithms. We have already seen that attempts
to use low quality MiSeq and HiSeq data causes computational problems
leading to the creation of open and closed reference clustering methods.
Unfortunately, comparative analyses showed that these methods fail
relative to \emph{de novo} clustering methods. More work is needed to
improve reference-based clustering methods so that larger datasets can
be analyzed without sacrificing quality. Finally, there are ongoing
controversies that need further exploration. These include the validity
and utility of amplicon sequence variants, the validity and impact of
removing low frequency sequences, and methods of identifying and
removing 16S rRNA gene sequences from contaminants. In the original
mothur paper, we commented that 16S rRNA gene sequencing and analysis is
very much like the Red Queen in Alice in Wonderland. Although some
disagreed with this analogy, we still feel it fits. The sequencing
technology and rapacious appetite of researchers continues to race on.
At the same time, bioinformatics tools must adapt to facilitate our
research. I am confident that mothur will be up to this exciting
challenge.
